Directionality in Kilo-ton Scale Scintillating Neutrino Detectors

### 2013-07-16 Direction.tex
Add most general references. --Lindley

\documentclass[aps,prc,twocolumn,groupedaddress,showpacs,amsmath,amssymb,floatfix,superscriptaddress]{revtex4}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{times}
\usepackage{textcomp}

@@

\section{Introduction}
Liquid scintillator based detectors are responsible for several of the critical measurements that have cemented our understanding of three-neutrino oscillations. These measurements include KamLAND's measurement reactor antineutrino oscillation at a distance of $\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar neutrino oscillation\cite{borexino} and most recently the short baseline reactor antineutrino experiments that measured oscillations due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo}, Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.  Scintillating neutrino detectors will continue to be important  for the next set of neutrino measurements from the determination of the neutrino mass hierarchy to elastic scattering measurements and sterile neutrino searches and for non-proliferation applications. Their scalability to large volumes makes them particularly attractive for neutrinoless double-beta ($0\nu\beta\beta$) decay searches, and currently one of the best limits for the $0\nu\beta\beta$ half-life comes from KamLAND-Zen.

@@

A good liquid scintillator will produce $\sim$10,000 photons isotropically per MeV of deposited energy. Although less abundant, if a particle is moving faster than the speed of light in the medium then Cerenkov light will be produced as well. This light is emitted in a cone pointed in the direction of the particles trajectory and with a continuous spectrum wavelengths weighted toward shorter wavelengths but extending well into the red,
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and $\beta$ is the velocity of the incoming particle. The Cerenkov light produced at wavelengths shorter than the absorption cutoff of the scintillator will be absorbed and re-emitted as isotropic light, but wavelengths longer than this cutoff will propagate across the detector retaining their directional information, roughly 60 photons per MeV assuming a 400~nm absorption cutoff. These undisturbed Cerenkov photons will have timing determined by the group velocity in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
These longer wavelengths Cerenkov photons arrive in principle before the scintillation light which is slowed by both the scintillation processes and the shorter wavelengths involved. Thus, with sufficient timing resolution and sensitivity to longer wavelength, it should be possible to separate the directional Cerenkov light and the isotropic scintillation light in time and then reconstruct the direction of the initial particle.

@@

Before we discuss the simulation results for different simulation settings in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Past neutrino experiments were not as interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation. We assume a rise time of 1~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (the group velocity is used in Geant4 for normal dispersion) is also wavelength dependent. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and did apply neither TTS nor TOF correction. The true time distribution of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are created about 0.5~ns earlier than PEs from scintillation light on average. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time. On the detector side, the time information of single photoelectrons is affected by the TTS of the photodetectors, a number which can be different by orders of magnitude depending on the detector type. The default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with LAPPDs \cite{tbd}. In fact, even significantly lower TTS numbers are realistic with this novel technology. Finally, the accuracy of the vertex reconstruction is reflected in broadening and distortions of the time spectrum (after the TOF correction has been applied). Another effect which is to be addressed in future work is the granularity of the photodetectors which adds to the uncertainty in the light path and thus the TOF correction.

>>>>

\documentclass[aps,prc,twocolumn,groupedaddress,showpacs,amsmath,amssymb,floatfix,superscriptaddress]{revtex4}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{times}
\usepackage{textcomp}
\usepackage{hyperref}

@@

\section{Introduction}
Liquid scintillator based detectors are responsible for several of the critical measurements that have cemented our understanding of three-neutrino oscillations. These measurements include KamLAND's measurement reactor antineutrino oscillation at a distance of $\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar neutrino oscillation\cite{borexino} and most recently the short baseline reactor antineutrino experiments that measured oscillations due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo}, Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.  Scintillating neutrino detectors will continue to be important  for the next set of neutrino measurements from the determination of the neutrino mass hierarchy\cite{juno,reno50} to elastic scattering measurements\cite{isodarscatt} and sterile neutrino searches\cite{isodar,nist} and for non-proliferation applications\cite{nucifer, songs}. Their scalability to large volumes makes them particularly attractive for neutrinoless double-beta ($0\nu\beta\beta$) decay searches, and currently one of the best limits for the $0\nu\beta\beta$ half-life comes from KamLAND-Zen\cite{KZ0nu}.

@@

A good liquid scintillator will produce $\sim$10,000 photons isotropically per MeV of deposited energy. Although less abundant, if a particle is moving faster than the speed of light in the medium then Cerenkov light will be produced as well. This light is emitted in a cone pointed in the direction of the particles trajectory and with a continuous spectrum wavelengths weighted toward shorter wavelengths but extending well into the red,
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and $\beta$ is the velocity of the incoming particle. The Cerenkov light produced at wavelengths shorter than the absorption cutoff of the scintillator will be absorbed and re-emitted as isotropic light, but wavelengths longer than this cutoff will propagate across the detector retaining their directional information, roughly 60 photons per MeV assuming a 400~nm absorption cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing determined by the group velocity in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
These longer wavelengths Cerenkov photons arrive in principle before the scintillation light which is slowed by both the scintillation processes and the shorter wavelengths involved. Thus, with sufficient timing resolution and sensitivity to longer wavelength, it should be possible to separate the directional Cerenkov light and the isotropic scintillation light in time and then reconstruct the direction of the initial particle.

@@

Before we discuss the simulation results for different simulation settings in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Past neutrino experiments were not as interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation. We assume a rise time of 1~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (the group velocity is used in Geant4 for normal dispersion) is also wavelength dependent. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and did apply neither TTS nor TOF correction. The true time distribution of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are created about 0.5~ns earlier than PEs from scintillation light on average. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time. On the detector side, the time information of single photoelectrons is affected by the TTS of the photodetectors, a number which can be different by orders of magnitude depending on the detector type. The default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with the large area picosecond photodetectors (LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors (HPDs)\cite{hpdThesis}. In fact, even significantly lower TTS numbers are realistic with the LAPPD. Finally, the accuracy of the vertex reconstruction is reflected in broadening and distortions of the time spectrum (after the TOF correction has been applied). Another effect which is to be addressed in future work is the granularity of the photodetectors which adds to the uncertainty in the light path and thus the TOF correction.
